---
title: "Data Visualization: Practice"
author: ""
date: ""
output: pdf_document
---

Let's create some graphs with `ggplot2`! You can use the power point and the `ggplot2` [cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf) for reference.

The data comes from the **2016 American National Election Study** (ANES). The ANES surveys a nationally representative sample of the U.S. public on voting, public opinion, and political participation.

The following code chunk includes code to load a practice dataset and to recode its variables. Run the code chunk in its entirety.

We are only looking at a small subset of variables:

- age (numeric)
- gender: 1 = male, 2 = female (categorical)
- partyID: 1 = Democrat, 2 = Republican, 3 = Independent (categorical)
- feelTrump: How would you rate Donald Trump? (0-100) (numeric)

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
rm(list=ls()) # clear out objects from workspace

#load data
library(readstata13)
anes16 <- read.dta13("http://jakebowers.org/PS531/anes_timeseries_2016.dta")

anes16 <- subset(anes16, select=c("V161267", "V161342", "V161155", "V161087"))

#change names of variables
install.packages("reshape")
library(reshape)
anes16 <- rename(anes16, c(V161267 ="age", V161342 = "gender", V161155 = "partyID", V161087 = "feelTrump"))

#recode variables
anes16$age <- with(anes16, ifelse(age < 0, NA, age))
anes16$feelTrump <- with(anes16, ifelse(feelTrump < 0, NA, feelTrump))
anes16$partyID <- with(anes16, ifelse(partyID < 1 | partyID > 4, NA, partyID))
anes16$partyID[anes16$partyID==1] <- "Democrat"
anes16$partyID[anes16$partyID==2] <- "Republican"
anes16$partyID[anes16$partyID==3] <- "Independent"
anes16$gender <- with(anes16, ifelse(gender < 0 | gender > 2, NA, gender))
anes16$gender[anes16$gender==1] <- "male"
anes16$gender[anes16$gender==2] <- "female"
anes16 <- anes16[complete.cases(anes16),]
```

1. What is the age distribution in the sample? (Hint: age is a continuous/numeric variable.) Add a title and label the axes of your plot.

```{r}

```


2. Does support for Trump (`feelTrump`) vary by gender? (Hint: one variable is numeric, the other categorical.) Add a title and label the axes of your plot.

```{r}

```